\section{Pattern Matching Compilation}
\label{sec:core}

the pattern matching algorithm of maranget~\cite{Maranget08} has been extended
to the setting of frank. here we need to be careful about order in the case of
open effect sets since forwarding is performed left-to-right. we do not
translate to core frank but instead compile to decision trees in the full
frank language (multi-handlers). how do we detect incomplete pattern
matching. we typecheck against the concrete set of effects and the type
signature. open sets allow an implicit forwarding clause. it is overridden if
the user provides their own forwarding clause. closed effect sets must be
fully specified. special wildcard pattern for computations desugars to perform
forwarding of unhandled commands -- this wildcard pattern is invalid in the
case of closed effect sets. the transformation results in additional rows for
the clause matrix and also requires a transformation of the expression on the
right hand side. it must execute the command and bind the result to be used in
a recursive call to the forwarding handler.

here is a rough description of what the algorithm does. we start with the
clause matrix of a multi-handler consisting of rows of source-level patterns
coupled with a mid-level computation (a clause). the algorithm compiles this
to a decision tree using an adapted version of maranget's approach (without
optimisations such as necessity). we convert the matrix to a column-first
representation. then pick the first column which has a pattern that is not a
wildcard. if the column is not the zeroth column then we perform a swap and
recurse (this will need to take account of forwarding). otherwise we compute
the type signature for the column (head constructors). we then compute the
type signature of the type for the column of patterns. comparing these will
determine whether or not we need a default/forwarding case. specifically, if
we have an open effect set. otherwise any discrepancy means that the user did
not specify all patterns and so this is a type error. constructing the
forwarding clauses shouldn't be too difficult. don't we just ignore the column
in question? or do we need to do some value threading? i think the answer
depends on the evaluation procedure.

one can think of the evaluation of a decision tree as column-first as opposed
to the row first approach that was previously taken in the evaluator. choosing
a column to compile should not alter the semantics of forwarding. so if we
cannot find a non-wildcard pattern to apply for the first column, what should
be done? i believe we should forward unhandled commands for that column. this
is column-first semantics. and indeed this will match the semantics of the
``row-first'' approach since even though the procedure checks by rows, we look
for a match on the first column, if no match, we check the next row, and so
on. the only difference is that the ``row-first'' approach would match on a
wildcard pattern (a thunk)...and in fact maybe this is what we should do...

picking a column that is not a wildcard. does not appear to be a valid
operation. because once we have picked a column. we are committed to
evaluating that particular argument. no wildcards. desugar patterns.
translate to core frank.

We take a fairly standard approach to compiling away pattern
matching. As we may match simultaneously against multiple
side-effecting computations, we must be somewhat careful about
order. Optionally, we can expose incomplete or ambiguous pattern
matching as concrete effects.

The target language of pattern matching compilation, Core Frank,
replaces multi-handlers with a combination of call-by-value functions,
case statements, and unary effect handlers.

The syntax of Core Frank is given in Figure~\ref{fig:core-syntax}.
%
\begin{figure}
Types
\begin{syntax}
\slab{values}       &U, V          &::=& D~\many{U} \mid \thunk{C} \mid X \\
\slab{computations} &C             &::=& R \mid V \to C \\
\slab{returners}    &R             &::=& \effbox{\sigs}V
\\[1ex]
\slab{quantifiers}  &Z             &::=& X \mid \varepsilon \\
\slab{polytypes}    &P             &::=& \forall \many{Z}.Q \\
\slab{thunks}       &Q             &::=& \thunk{C}
\\[1ex]
\slab{signatures}   &\sig~\many{X} &::=& \cdot \mid c : U(\many{V}), \sig~\many{X} \\
\slab{effects}      &\sigs         &::=&
  \emptyset \mid \sigs, \sig~\many{V} \mid \varepsilon
\\[1ex]
\slab{type environments}
                    &\Gamma        &::=& \cdot \mid \Gamma, x:V \mid f:P \\
\end{syntax}

Terms
\begin{syntax}
\slab{inferable values}&u       &::=& x \mid f \mid c \mid d \\
\slab{checkable values}&v       &::=& u \mid k~\many{v} \mid \thunk{e}
\\[1ex]
\slab{inferable computations}&d &::=& \force{u} \mid d~v \\
%                               \mid  \key{let}~x=d~\key{in}~d' \\
                             &&\mid& \key{letrec}~\many{f : P = e}~\key{in}~d \\
\slab{checkable computations} &e  &::= & v \mid \lambda x.e \\
   &&\mid& \key{case}~u~\key{of}~
             (k~\many{x_k} \mapsto e_k)_k \\
   &&\mid& \key{handle}~d~\key{with}~
             \bstack
             (\handle{c~\many{x_c}}{g_c} \mapsto e_c)_c \\
             \medvert~ x        \mapsto e \\
             \estack \\
\end{syntax}
\caption{Core Frank Syntax}
\label{fig:core-syntax}
\end{figure}
%
The Core Frank typing rules are given in Figure~\ref{fig:core-typing}.
%
\begin{figure*}[float]
$\boxed{\effs{C}{\sigs}}$

\begin{mathpar}
\inferrule
  { }
  {\effs{\effbox{\sigs}V}{\sigs}}

\inferrule
  {\effs{C}{\sigs}}
  {\effs{V \to C}{\sigs}}
\end{mathpar}

$\boxed{\makes{\Gamma}{\sigs}{u}{V}}$
\begin{mathpar}
\inferrule
  {x:V \in \Gamma}
  {\makesgs{x}{V}}

%% \inferrule
%%   {\dom(\theta) = \set{\varepsilon} \cup \FV(V) \backslash (\FV(\Gamma) \cup \FV(\sigs)) \\
%%    \theta(\varepsilon) = \sigs}
%%   {\makes{\Gamma, x:V}{\sigs}{x}{\theta(V)}}

\inferrule
  {f:\forall \many{Z}.Q \in \Gamma \\
   \dom(\theta) = \set{\many{Z}} \\
   \varepsilon \in \dom(\theta) \implies \theta(\varepsilon) = \sigs}
  {\makesgs{f}{\theta(Q)}}

\inferrule
  {c : U(\many{V}) \in \sigs}
  {\makesgs{c}{\thunk{\many{V} \to \effbox{\sigs}U}}}

\inferrule
  {\cangs{d}{\effbox{\sigs}V}}
  {\makesgs{d}{V}}
\end{mathpar}

$\boxed{\has{\Gamma}{\sigs}{V}{v}}$

\begin{mathpar}
\inferrule
  {\makesgs{u}{U} \\ U = V}
  {\hasgs{V}{u}}

\inferrule
  {(\hasgs{V_i}{v_i})_i \\
   k~\many{V} \in D~\many{U}}
  {\hasgs{D~\many{U}}{k~\many{v}}}

\inferrule
  {\does{\Gamma}{C}{e}}
  {\hasgs{\thunk{C}}{\thunk{e}}}
\end{mathpar}

$\boxed{\can{\Gamma}{\sigs}{d}{C}}$

\begin{mathpar}
\inferrule
  {\makesgs{u}{\thunk{C}} \\ \effs{C}{\sigs}}
  {\cangs{\force{u}}{C}}

\inferrule
  {\cangs{d}{V \to C} \\
   \has{\Gamma}{\sigs}{V}{v}}
  {\can{\Gamma}{\sigs}{d~v}{C}}

%% \inferrule
%%   {c : U(\many{V}) \in \sigs \\
%%    (\hasgs{V_i}{v_i})_i}
%%   {\cangs{c~\many{v}}{\effbox{\sigs}U}}

%% \inferrule
%%   {\cangs{d}{V} \\
%%    \can{\Gamma, x:V}{\sigs}{d'}{C}}
%%   {\cangs{\key{let}~x=d~\key{in}~d'}{C}}

\inferrule
  {(\does{\Gamma, \many{f : \forall \many{Z}.Q}}{e_i}{Q_i})_i \\
   \can{\Gamma, \many{f : \forall \many{Z}.Q}}{\sigs}{d}{C}}
  {\cangs{\key{letrec}~\many{f : \forall \many{Z}.Q = e}~\key{in}~d}{C}}
\end{mathpar}

$\boxed{\doesg{C}{e}}$

\begin{mathpar}
\inferrule
  {\hasgs{V}{v}}
  {\doesg{\effbox{\sigs}V}{v}}

\inferrule
  {\does{\Gamma, x:V}{C}{e}}
  {\doesg{V \to C}{\lambda x.e}}

\inferrule
  {\makesgs{u}{D~\many{U}} \\
   \\ \effs{C}{\sigs} \\
   (\does{\Gamma, \many{x_k}:\many{V}}{C}{e_k})_{k~\many{V} \in D~\many{U}}}
  {\doesg{C}{\key{case}~ u ~\key{of}~
               (k~\many{x_k} \mapsto e_k)_k}}

\inferrule
  {\sigs' \text{ closed} \\
   \effs{C}{\sigs} \\
   \can{\Gamma}{\sigs \oplus \sigs'}{d}{\effbox{\sigs \oplus \sigs'}{V}} \\
   (\does{\Gamma, \many{x_c}:\many{V}, g_c:\thunk{U \to \effbox{\sigs \oplus \sigs'}{V}}}
         {C}{e_c})_{c : U(\many{V}) \in \sigs'} \\
   \does{\Gamma, x:V}{C}{e}}
  {\does{\Gamma}
         {C}{\key{handle}~ d ~\key{with}~
               (\handle{c~\many{x_c}}{g_c} \mapsto e_c)_c \medvert
                x \mapsto e}}
\end{mathpar}

\caption{Core Frank Typing Rules}
\label{fig:core-typing}
\end{figure*}


%% The syntax of Core Frank is given in Figure~\ref{fig:core-syntax}.

%% \begin{figure}
%% Types
%% \begin{syntax}
%% \slab{values}       &U, V   &::=& D~\many{U} \mid  \effbox{\sigs}\thunk{C} \mid X \\
%% \slab{computations} &C      &::=& \rt{V} \mid V \to C
%% \\[1ex]
%% \slab{polytypes}    &P      &::=& \forall \varepsilon \many{X}.Q \\
%% \slab{thunks}       &Q      &::=& \effbox{\sigs}\thunk{C}
%% \\[1ex]
%% \slab{signatures}   &\sig~\many{X} &::=& \cdot \mid c : U(\many{V}), \sig~\many{X} \\
%% \slab{effects}      &\sigs  &::=&
%%   \emptyset \mid \sigs, \sig~\many{V} \mid \varepsilon
%% \\[1ex]
%% \slab{type environments}     &\Gamma &::=& \cdot \mid \Gamma, x:V \mid f:P \\
%% \end{syntax}

%% Terms
%% \begin{syntax}
%% \slab{inferable values}       &u  &::= & x \mid f \mid d                           \\
%% \slab{checkable values}       &v  &::= & u \mid k~\many{v} \mid \thunk{e}
%% \\[1ex]
%% \slab{inferable computations} &d  &::= & \force{u} \mid d~v \mid c~\many{v}
%%                                    \mid  \key{let}~x=d~\key{in}~d' \\
%%                               &   &\mid& \key{letrec}~\many{f : P = e}~\key{in}~d \\
%% \slab{checkable computations} &e  &::= & v \mid \lambda x.e \\
%%   &&\mid& \key{case}~u~\key{of}~
%%             (k~\many{x_k} \mapsto e_k)_k \\
%%   &&\mid& \key{handle} ~d~ \key{with}
%%             \bstack
%%             ~~(\handle{c~\many{x_c}}{g_c} \mapsto e_c)_c \\
%%             \medvert~ x        \mapsto e \\
%%             \estack \\
%% \end{syntax}
%% \caption{Core Frank Syntax}
%% \label{fig:core-syntax}
%% \end{figure}


%% A superficial change to the types is that rather than annotating value
%% returning computations with effects, we shift such labels to the thunk
%% containing the computation. The former design seems more convenient to
%% program with, which is why we adopt it in the source language. The
%% latter design leads to a slightly more uniform presentation of the
%% type rules (the checking judgement for computations now has the same
%% shape as the checking judgement for values, and we no longer need the
%% $\judgeword{does}$ judgement).
%
%% The typing rules of Core Frank are given in
%% Figure~\ref{fig:core-syntax}.

%%%%% Core Frank

%% \begin{figure*}[float]
%% $\boxed{\makes{\Gamma}{\sigs}{u}{V}}$
%% \begin{mathpar}
%% \inferrule
%%   {x:V \in \Gamma}
%%   {\makesgs{x}{V}}

%% %% \inferrule
%% %%   {\dom(\theta) = \set{\varepsilon} \cup \FV(V) \backslash (\FV(\Gamma) \cup \FV(\sigs)) \\
%% %%    \theta(\varepsilon) = \sigs}
%% %%   {\makes{\Gamma, x:V}{\sigs}{x}{\theta(V)}}

%% \inferrule
%%   {f:\forall \varepsilon \many{X}.V \in \Gamma \\
%%    \dom(\theta) = \set{\varepsilon, \many{X}} \\
%%    \theta(\varepsilon) = \sigs}
%%   {\makesgs{f}{\theta(V)}}

%% \inferrule
%%   {\cangs{d}{\rt{V}}}
%%   {\makesgs{d}{V}}
%% \end{mathpar}

%% $\boxed{\has{\Gamma}{\sigs}{V}{v}}$

%% \begin{mathpar}
%% \inferrule
%%   {\makesgs{u}{V}}
%%   {\hasgs{V}{u}}

%% \inferrule
%%   {(\hasgs{V_i}{v_i})_i \\
%%    k~\many{V} \in D~\many{U}}
%%   {\hasgs{D~\many{U}}{k~\many{v}}}

%% \inferrule
%%   {\cdoes{\Gamma}{\sigs'}{C}{e}}
%%   {\hasgs{\effbox{\sigs'}\thunk{C}}{\thunk{e}}}
%% \end{mathpar}

%% $\boxed{\can{\Gamma}{\sigs}{d}{C}}$

%% \begin{mathpar}
%% \inferrule
%%   {\makesgs{u}{\effbox{\sigs}\thunk{C}}}
%%   {\cangs{\force{u}}{C}}

%% \inferrule
%%   {\cangs{d}{V \to C} \\
%%    \has{\Gamma}{\sigs}{V}{v}}
%%   {\cangs{d~v}{C}}

%% \inferrule
%%   {c : U(\many{V}) \in \sigs \\
%%    (\hasgs{V_i}{v_i})_i}
%%   {\cangs{c~\many{v}}{\rt{U}}}

%% \inferrule
%%   {\cangs{d}{\effbox{\sigs}V} \\
%%    \can{\Gamma, x:V}{\sigs}{d'}{C}}
%%   {\cangs{\key{let}~x=d~\key{in}~d'}{C}}

%% \inferrule
%%   {(\does{\Gamma, \many{f : \forall \varepsilon \many{X}.Q}}{e_i}{Q_i})_i \\
%%    \can{\Gamma, \many{f : \forall \varepsilon \many{X}.Q}}{\sigs}{d}{C}}
%%   {\cangs{\key{letrec}~\many{f : \forall \varepsilon \many{X}.Q = e}~\key{in}~d}{C}}
%% \end{mathpar}

%% $\boxed{\cdoesgs{C}{e}}$

%% \begin{mathpar}
%% \inferrule
%%   {\hasgs{V}{v}}
%%   {\cdoesgs{\rt{V}}{v}}

%% \inferrule
%%   {\cdoes{\Gamma, x:V}{\sigs}{C}{e}}
%%   {\cdoesgs{V \to C}{\lambda x.e}}

%% \inferrule
%%   {\makesgs{u}{D~\many{U}} \\
%%    (\cdoes{\Gamma, \many{x_k}:\many{V}}{\sigs}{C}{e_k})_{k~\many{V} \in D~\many{U}}}
%%   {\cdoesgs{C}{\key{case}~ u ~\key{of}~
%%                (k~\many{x_k} \mapsto e_k)_k}}

%% \inferrule
%%   {\sigs' \text{ closed} \\
%%    \can{\Gamma}{\sigs \oplus \sigs'}{d}{\rt{V}} \\
%%    (\cdoes{\Gamma, \many{x_c}:\many{V}, g_c:\effbox{\sigs \oplus \sigs'}\thunk{U \to \rt{V}}}
%%          {\sigs}{C}{e_c})_{c : U(\many{V}) \in \sigs'} \\
%%    \cdoes{\Gamma, x:V}{\sigs}{C}{e}}
%%   {\cdoes{\Gamma}{\sigs}
%%          {C}{\key{handle}~ d ~\key{with}~
%%                (\handle{c~\many{x_c}}{g_c} \mapsto e_c)_c \medvert
%%                 x \mapsto e}}
%% \end{mathpar}

%% \caption{Core Frank Typing Rules}
%% \label{fig:core-typing}
%% \end{figure*}

Multi-handlers in Frank become curried functions over suspended
computations in Core Frank.
%
Shallow pattern matching on a single request becomes unary effect
handling. Shallow pattern matching on a datatype value becomes case
analysis. Nested pattern matching on multiple computations is realised
as a pattern matching tree constructed from handlers and case
statements. 

Standard algorithms for pattern matching compilation
apply~(e.g. \cite{Augustsson85} or \cite{Maranget08}). we commit to the
algorithm set out by Maranget in the implementation but adapted to handle the
value/computation distinction. Rather than committing to a particular one, we
outline how a pattern matching compiler fits into our setting, what input it
takes, and what kind of output it must produce.

\begin{sloppypar}
Given a Frank expression $\thunk{e}$ such that $\doesg{R_1 \dots R_n
  \to R}{e}$
%
we compile it to an equivalent Core Frank expression $\thunk{\pc{e}}$.
%
First we expand all of the clauses in $e$ to yield an $n$ column
pattern matrix. For instance, suppose the arguments have types
$\effbox{\var{Send~Char}, \var{Abort}}\var{Unit}$ and
$\effbox{\var{Receive~Char}}$, and we have the following clauses:
\end{sloppypar}
\[
\ba{@{}l@{~}l@{~}l@{}}
  (\handle{\var{send}~x}{s}) & (\handle{\var{receive}}{r}) & \mapsto e_1 \\
  (\handle{\var{send}~x}{s}) & z & \mapsto e_2 \\
  (\handle{\var{abort}}{s})  & e_3 \\
  \var{unit} & e_4 \\
\ea
\]
%
then this becomes:
%
\[
\ba{@{}l@{\quad}l@{~\mapsto~}l@{}}
  (\handle{\var{send}~x}{s}) & (\handle{\var{receive}}{r}) & e_1 \\
  (\handle{\var{send}~x}{s}) & z                           & e_2 \\
  (\handle{\var{abort}}{s})  & \force{y} & \var{bind}~\force{y}~e_3 \\
  \var{unit}                 & \force{y} & \var{bind}~\force{y}~e_4 \\
\ea
\]
%
where $\var{bind}$ is defined in Section~\ref{sec:examples}. The
missing patterns have been inserted as thunk patterns, which match any
computation. Invoking $\var{bind}$ allows us to forward the
computation bound by the thunk pattern to the existing continuation.
%
Next we generate a vector of fresh variables, one for each argument.
\[
\ba{@{}l@{~}l@{}}
  x_0 & x_1
\ea
\]
The goal of pattern matching compilation is to generate a pattern
matching tree that matches the variable vector against all of the
patterns in the pattern matrix in the correct order.

In Frank, pattern matching trees $M$ are built up from leaves, case
analysis, and handlers.
%% \begin{equations}
%% M &::= & \many{e} \\
%%   &\mid& \key{case}~(x:D~\many{U})~\key{of}~(k~\many{x_k} \mapsto M_k)_{k \in D} \\
%%   &\mid& \key{handle}~(\force{x}:\effbox{\sigs}V)~\key{with}~
%%           (\handle{c~\many{x_c}}{g_c} \mapsto M_c)_{c \in \sigs} \medvert
%%           x        \mapsto M
%% \end{equations}%
\begin{equations}
M &::= & \many{e} \\
  &\mid& \key{case}~x~\key{of}~(k~\many{x_k} \mapsto M_k)_{k \in D} \\
  &\mid& \key{handle}~\force{x}~\key{with}~
          (\handle{c~\many{x_c}}{g_c} \mapsto M_c)_{c \in \sigs} \medvert
          x        \mapsto M
\end{equations}%
The leaves consist of a sequence of checkable computation
expressions. Each element corresponds to one way of matching all of
the patterns. If there exists a leaf with no elements, then the
pattern matching is incomplete; if there exists a leaf with multiple
elements, then the pattern matching is ambiguous.
%
Our default strategy (as indicated by the thunk introduction rule) is
to class incomplete pattern matching as a type error, and to keep only
the first element in the case of ambiguous pattern matching.
%
Our example generates the following pattern matching tree:
\[
M = \bstack
    \key{handle}~\force{x_0}~\key{with} \\
    \quad \handle{\var{send}~x}{s} \mapsto \\
    \quad\quad \key{handle}~\force{x_1\!}~\key{with} \\
    \quad\quad\quad \begin{clauses}
                    \handle{\var{receive}}{r} &\mapsto& e_1 \\
                    z                         &\mapsto& e_2 \\
                    \end{clauses} \\
    \quad \handle{abort}{s} \mapsto \\
    \quad\quad \key{handle}~\force{x_1\!}~\key{with} \\
    \quad\quad\quad \begin{clauses}
                    \handle{receive}{r} &\mapsto&
                      \var{bind}~(\force{r}~\force{receive})~e_3 \\
                    z &\mapsto& \var{bind}~z~e_3 \\
                    \end{clauses} \\
    \quad  y \mapsto \\
    \quad\quad  \key{case}~y~\key{of} \\
    \quad\quad\quad  \var{unit} \mapsto \key{handle}~\force{x_1\!}~\key{with} \\
    \quad\quad\quad\quad \begin{clauses}
                         \handle{receive}{r} &\mapsto&
                            \var{bind}~(\force{r}~\force{receive})~e_4 \\
                         z &\mapsto& \var{bind}~z~e_4 \\
                         \end{clauses} \\
\estack
\]
Each thunk pattern has been expanded out to explicitly list all of the
cases according to its type. We obtain the corresponding Core Frank
code by abstracting over the fresh variables.
%
\[
\lambda x_0~x_1.M
\]
Some pattern matching operations reorder columns as an
optimisation. Column reordering is not in general a valid optimisation
in Frank. This is because commands in the ambient effects, but not in
the argument effects, are implicitly forwarded, and the order in which
they are forwarded is left-to-right. (The forwarding behaviour is made
precise in the Section~\ref{sec:semantics}.)

Of course, because Core Frank takes values as arguments whereas Frank
takes computations, each argument must be wrapped in a thunk
constructor.
%
The type translation is given simply by the homomorphic extension of the
following equation on function types:
\begin{equations}
\pc{\effbox{\sigs}V} \to \pc{C} &=& \thunk{\effbox{\pc{\sigs}}\pc{V}} \to \pc{C} \\
\end{equations}

A correct pattern matching translation $\pc{-}$ from Frank to Core
Frank should be type preserving.
\begin{itemize}
\item If $\makesgs{u}{V}$ then $\makes{\pc{\Gamma}}{\pc{\sigs}}{\pc{u}}{\pc{V}}$.
\item If $\hasgs{V}{v}$   then $\has{\pc{\Gamma}}{\pc{\sigs}}{\pc{V}}{\pc{v}}$.
\item If $\cangs{d}{C}$   then $\can{\pc{\Gamma}}{\pc{\sigs}}{\pc{d}}{\pc{C}}$.
%%\item If $\doesg{C}{e}$  then $\does{\pc{\Gamma}}{\pc{C}}{\pc{e}}$.
\end{itemize}

\subsection{Incomplete and Ambiguous Pattern Matching as Effects}

As an extension to Frank, we might allow incomplete and ambiguous
pattern matching. The former may be permitted if the ambient effects
contain the $\var{Abort}$ signature, in which case incomplete patterns
are translated into the $\var{abort}:\var{Zero}$ command, which can
then be handled however the programmer wishes. Similarly, we can define
a $\var{choice}:X(X, X)$ command, in order to allow ambiguous pattern
matches to be handled by the programmer.

%% \begin{equations}
%% \pc{D~\many{U}) &=& D~\pc{\many{U}} \\
%% \pc{\thunk{C}}  &=& \thunk{\pc{C}} \\
%% \pc{X} &=& X
%% \\[1ex]
%% \pc{\effbox{\sigs}V} &=& \effbox{\pc{\sigs}}\pc{V} \\
%% \pc{\effbox{\sigs}V} \to \pc{C} &=& \thunk{\effbox{\pc{\sigs}}\pc{V}} \to \pc{C} \\
%% \\[1ex]
%% \pc{\emptyset} &=& \emptyset \\
%% \pc{\sigs, S~\many{V}} &=& \pc{\sigs}, S~\pc{\many{V} \\
%% \pc{\varepsilon} &=& \varepsilon \\
%% \\[1ex]
%% \pc{\cdot} &=& \cdot \\
%% \pc{c:U(\many{V}), S~\many{X}} &=& 
%% \end{equations}

%%  LocalWords:  Lindley Conor Strathclyde Plotkin Pretnar's Multi et
%%  LocalWords:  effectful modularity polymorphism Kammar al Pretnar
%%  LocalWords:  Hindley Milner Oury SML Haskell monadic Kiselyov DSL
%%  LocalWords:  Idris Brady's equational multi unary al's datatypes
%%  LocalWords:  datatype Vn typecheck Ri catter polytypes inferable
%%  LocalWords:  checkable letrec monomorphic reifying desugars ret
%%  LocalWords:  homomorphic monothunks reified pre rewrapped monads
%%  LocalWords:  denotationally Swierstra Visscher Filinski Daan Koka
%%  LocalWords:  Filinski's Leijen's Swamy monad
